Exact analytic solution for slope flows with spatially varying eddy viscosity and diffusivity

### 2015-11-03 paper_BLM.tex
latest version

\section{Introduction}
Natural convection of turbulent stratified flows over sloping surfaces is ubiquitous in nature and engineering, and is of interest not only as a fundamental problem in itself, but also because of the important role it plays over a broad range of scales and applications. 
On a local scale, for instance, it governs local weather conditions, affecting atmospheric transport processes and dispersion of scalars such as heat and humidity \citep{Monti2002, Nylen2004b}. 
On a regional scale, it drives the development of deep-convective circulations \citep{Egger1985, Parish1998} and is responsible for intense cyclonic vorticity in the middle and upper troposphere \citep{Parish1992}.
Further, recent experimental field campaigns \citep{Chu1987, Oerlemans1993} have shown that persistent katabatic winds characterize the atmospheric boundary layer over snow-ice surfaces and glaciers, and therefore an accurate characterization of such flows is an essential component of understanding and modeling of the weather and climate.
However, the complex dynamics (e.g. occurrence of intermittency, waves, Kelvin-Helmholtz instabilities and low-level jets) and the lack of a satisfactory similarity theory for such flows \citep{Nadeau2012} pose a heavy burden in terms of computational requirements for numerical modellers; in most cases the required resolution is prohibitively costly \citep{Fedorovich2009, Fedorovich2009d, Burkholder2011a}.
Because of this, conceptual models are still of great interest, and represent a valid tool for the characterization of such systems. 
A cornerstone in the understanding of natural convection of stratified fluid along a sloping surface is represented by the classic Prandtl analytic model \citep{Prandtl}, and its more recent extensions/generalizations to include the effects of Coriolis force \citep{Gutman1964}, external winds \citep{Lykosov_1972}, surface heterogeneity \citep{Shapiro2007, Oldroyd2014}.
The Prandtl model approximates the atmosphere in a Boussinesq sense and describes a steady flow over a thermally perturbed unbounded planar sloping surface that lies within a stratified environment, given by the following system of equations:
%
\begin{linenomath*}
\begin{eqnarray}
	N^2 u(z) \sin{\alpha} & = & d( K_H db/dz )/dz,
	\label{eqn1u} \\
	%
	b(z) \sin{\alpha} & = & -d( K_M du/dz )/dz,
	\label{eqn1b}
\end{eqnarray}
\end{linenomath*}
%
where $z$ denotes the normal-to-slope coordinate directions, $u$ is the downslope velocity, $b$ is buoyancy, $N$ is the buoyancy frequency characterizing the system (related to the background stratification), $\alpha$ is the slope angle and $K_M$ and $K_H$ denote the eddy viscosity and diffusivity (an eddy viscosity/diffusivity model has been used to parametrize turbulent fluxes of momentum and buoyancy).
Equations are defined in $z \in \ [0,\infty)$ with boundary conditions $u(0)=0$, $u(z \rightarrow \infty) = 0$, $b(0)=b_s$ and $b(z \rightarrow \infty) = 0$ ($b_s > 0$ for upslope flows, whereas $b_s<0$ for downslope flows).

@@

Assuming eddy diffusivities in line with the classic O'Brien's type \citep{O'Brien1970}, viz. $K(z) = A (z + \epsilon)(z-\zeta)^2$, where $\epsilon>0$ and $\zeta>H$ are parameters that we introduce in order to shift the zeroes of $K$ outside the interval of integration $[0,H]$, assuming $Pr=1$ (i.e. $K_M=K_H=K$) and assigning a length, velocity and buoyancy scale $L= (N \sin{\alpha})/ A$, $U=b_s N^{-1}$ and $B=b_s$ respectively, Eqs. \ref{eqn1u} and \ref{eqn1b} reduce to:

@@

	\overline{u} & = d( \overline{K} d\overline{b}/d\overline{z} )/d\overline{z}, && \mathrm{in}  \ [0,\overline{H}]

@@

	\overline{b} & = -d( \overline{K} d\overline{u}/d\overline{z} )/d\overline{z}, &&\mathrm{in}  \ [0,\overline{H}]

@@

where $\overline{K} = (\overline{z} + \overline{\epsilon})(\overline{z}-\overline{\zeta})^2$ is the normalized eddy viscosity/diffusivity, $\overline{z}=zL^{-1}$, $\overline{u} = uU^{-1}$, $\overline{b}=b B^{-1}$, with boundary conditions $\overline{u}(0)=0$, $\overline{u}(\overline{H}) = 0$, $\overline{b}(0)=\pm 1$ and $\overline{b}(\overline{H}) = 0$.

@@

	i \cdot f =  d(  K df/dz )/dz \hspace{1cm} \mathrm{in}  \ [0,H],

@@

with boundary conditions $f(0)=\pm 1$ and $f(H) = 0$.

@@

where $(\cdot)' = \frac{\partial (\cdot)}{\partial z}$, $P(z)=(K)'/K$ and $Q(z)=1/(i K)$. Second, it is an easy computation to show that rewriting Eq. \ref{can_form} for $y = \frac{z + \epsilon}{H + \epsilon + \zeta}

@@

where $(\cdot)' = \frac{\partial (\cdot)}{\partial y}$, $\widetilde{P}(y)=\gamma'(y)/\gamma(y)$ and $\widetilde{Q}(y)=1/( i \cdot (H+\epsilon+\zeta) \gamma(y))$, with $\gamma(y)=y(y-1)^2$.

@@

	x^2+x+1/( i (H+ \epsilon + \zeta))=0.

@@

The model assumes $Pr = 1$, i.e. $K_M = K_H = K$, although in reality they are slightly different. This assumption is justified based on the degree of accuracy of the considered study. 
Further, as stated in the preceeding section, we assume eddy diffusivities in line with the classic O'Brien's type \citep{O'Brien1970}, viz. $K = A(z + \epsilon)(z-\zeta)^2$, where $\epsilon>0$ can be tuned to provide the desired value of eddy viscosity / diffusivity at the wall, whereas $\zeta > H$ controls both shape and magnitude of $K$.
The O'Brien model has often been used in studies of stable boundary layers, see for instance \citet{pielke1984mesoscale} and \citet{Stull1988}, and a generalized O'Brien model was also recently adopted in \citet{grisogono2001katabatic} to study slope flows.
The properties of $K(z)$ in slope flows were thorougly justified in \citet{Grisogono2002};  we here recall the main criteria that $K(z)$ should honor:
%
\begin{enumerate}
\item realizability condition: $K(z) \ge 0$;
\item $K(z << \infty) = 0$;
\item $\max(K(z))$, reached at $z=h$, must lie within the dynamic and thermal boundary layers;
\item $h > \max(2 z_j, z_{inv})$;
\end{enumerate}
%
where $z_j$ is the height of the LLJ, and $z_{inv}$ is the height of the surface inversion layer (in general we have $z_{inv} > z_j$).
%
In addition, assuming Monin-Obukhov similarity theory to hold for the near wall regions (which is itself a subject of great debate), one would ideally specify $K(z)$ in order to satisfy:
%
\begin{linenomath*}
\begin{subequations}
\begin{align}
   K(z) & = \kappa u_* z/[1+\Phi(z)] & z \in [z_0,z_s], \label{MOST1}\\
   K(z) & = f(u_*,z) \ | \ d^2f/dz^2 < 0 & z \in (z_s,H], \label{MOST2}
\end{align}
\end{subequations}
\end{linenomath*}
%
where $z_s$ is the upper boundary of the constant-flux layer, $\Phi(z)$ is a stability correction function, $\kappa$ is the Von K\'arm\'an constant, and $u_* = \sqrt{K(du/dz)/\rho}$ is the so-called friction velocity.
However, in the current settings, we are not able to satisfy Eqs. \ref{MOST1} and \ref{MOST2} directly, since the specification of $K$ has to be done a-priori, and $u_*$ depends on the solution. 
This is the main weakness of linear approaches. The only way to enforce Eqs. \ref{MOST1} and \ref{MOST2} would be to apply a procedure in which $K(z)$ is iteratively updated to match the MO similarity relations.
Based on recent direct (DNS) and large-eddy simulation (LES) results (not shown) we make the additional assumption that $h$ is located in the neighborhood of the first zero of the $u$ velocity profile, well above $z_j$ but within the dynamic and thermal boundary layers ($h$ would be at the lower boundary of the return flow).
Knowledge of $K(z_0),dK(z_0)/dz$ and $h$ allows to univocally define $K(z)= A(z + \epsilon)(z-\zeta)^2$ one can therefore specify the surface value and its derivative $K(0), dK(0)/dz$ and 
Under such restrictive (but realistic) assumption it is possible to show that the $H$ parameter in

Despite the restrictive form of the model parameter, which might limit the applicability of the solution in certain contexts, the model

@@

      \includegraphics[width= 70.0mm]{./u-b_log}
      \caption{Comparison of the new analytic solution (A1) against the constant-$K$ (A2), the WKB (A3) and a numerical $1D$ solution (N1). Analytic profiles of normalized velocity $(u^+)$ are denoted with dot-dashed lines whereas analytic profiles of normalized buoyancy $(b^+)$ are denoted by dashed lines. Here $K^+ = A^+ \cdot (z^++\epsilon^+)(z^+-H^+-\epsilon^+)^2$ where $A^+=6.75 e-4$, $\epsilon^+ =1.5 \times 10^{-3}$ and $H^+ = 10$. A corresponding dimensional profile is characterized by $\max{(K)} = 0.1 \ \mathrm{m^2s^{-1}}$ and $K(z=0) = 10^{-5} \ \mathrm{m^2s^{-1}}$, in agreement with common atmospheric values.}

@@

In Fig. \ref{fig1} we compare the new analytic solution (A1 in the following) against the constant-$K$ (A2), the WKB (A3) and a numerical solution (N1).

@@

In comparison to its analytical counterparts, A1 shows a remarkably strong inversion in the near surface regions $(z^+ <1)$, suggesting overdiffusive behaviour of both A2 and A3, which are incapable of providing such strong buoyancy gradients. 

>>>>

\section{Introduction}
Natural convection of turbulent stratified flows over sloping surfaces is ubiquitous in nature and engineering, and is of interest not only as a fundamental problem in itself, but also because of the important role it plays over a broad range of scales and applications. 
On a local scale, for instance, it governs local weather conditions, affecting atmospheric transport processes and dispersion of scalars such as heat and humidity \citep{Monti2002, Nylen2004b}. 
On a regional scale, it drives the development of deep-convective circulations \citep{Egger1985, Parish1998} and is responsible for intense cyclonic vorticity in the middle and upper troposphere \citep{Parish1992}.
Further, recent experimental field campaigns \citep{Chu1987, Oerlemans1993} have shown that persistent katabatic winds characterize the atmospheric boundary layer over snow-ice surfaces and glaciers, and therefore an accurate characterization of such flows is an essential component of understanding and modeling of the weather and climate.
However, the complex dynamics (e.g. occurrence of intermittency, waves, Kelvin-Helmholtz instabilities and low-level jets) and the lack of a satisfactory similarity theory for such flows \citep{Nadeau2012} pose a heavy burden in terms of computational requirements for numerical modellers; in most cases the required resolution is prohibitively costly \citep{Fedorovich2009, Fedorovich2009d, Burkholder2011a}.
Because of this, conceptual models are still of great interest, and represent a valid tool for the characterization of such systems. 
A cornerstone in the understanding of natural convection of stratified fluid along a sloping surface is represented by the classic Prandtl analytic model \citep{Prandtl}, and its more recent extensions/generalizations to include the effects of Coriolis force \citep{Gutman1964}, external winds \citep{Lykosov_1972}, surface heterogeneity \citep{Shapiro2007, Oldroyd2014}.
The Prandtl model approximates the atmosphere in a Boussinesq sense and describes a steady flow over a thermally perturbed unbounded planar sloping surface that lies within a stratified environment, given by the following system of equations:
%
\begin{linenomath*}
\begin{eqnarray}
	N^2 u(z) \sin{\alpha} & = & d( K_H db/dz )/dz,
	\label{eqn1u} \\
	%
	b(z) \sin{\alpha} & = & -d( K_M du/dz )/dz,
	\label{eqn1b}
\end{eqnarray}
\end{linenomath*}
%
where $z$ denotes the normal-to-slope coordinate directions, $u$ is the downslope velocity, $b$ is buoyancy, $N$ is the buoyancy frequency characterizing the system (related to the background stratification), $\alpha$ is the slope angle and $K_M$ and $K_H$ denote the eddy viscosity and diffusivity (an eddy viscosity/diffusivity model has been used to parametrize turbulent fluxes of momentum and buoyancy).
Equations are defined in $z \in \ [z_0,\infty)$ with boundary conditions $u(z_0)=0$, $u(z \rightarrow \infty) = 0$, $b(z_0)=b_s$ and $b(z \rightarrow \infty) = 0$ ($b_s > 0$ for upslope flows, whereas $b_s<0$ for downslope flows).

@@

Assuming eddy diffusivities in line with the classic O'Brien's type \citep{O'Brien1970}, viz. $K(z) = A z(z-\zeta)^2$, where $\zeta>H$, $Pr=1$ (i.e. $K_M=K_H=K$) and assigning a length, velocity and buoyancy scale $L= (N \sin{\alpha})/ A$, $U=b_s N^{-1}$ and $B=b_s$ respectively, Eqs. \ref{eqn1u} and \ref{eqn1b} reduce to:

@@

	\overline{u} & = d( \overline{K} d\overline{b}/d\overline{z} )/d\overline{z}, && \mathrm{in}  \ [z_0,\overline{H}]

@@

	\overline{b} & = -d( \overline{K} d\overline{u}/d\overline{z} )/d\overline{z}, &&\mathrm{in}  \ [z_0,\overline{H}]

@@

where $\overline{K} = \overline{z}(\overline{z}-\overline{\zeta})^2$ is the normalized eddy viscosity/diffusivity, $\overline{z}=zL^{-1}$, $\overline{u} = uU^{-1}$, $\overline{b}=b B^{-1}$, with boundary conditions $\overline{u}(z_0)=0$, $\overline{u}(\overline{H}) = 0$, $\overline{b}(z_0)=\pm 1$ and $\overline{b}(\overline{H}) = 0$.

@@

	i \cdot f =  d(  K df/dz )/dz \hspace{1cm} \mathrm{in}  \ [z_0,H],

@@

with boundary conditions $f(z_0)=\pm 1$ and $f(H) = 0$.

@@

where $(\cdot)' = \frac{\partial (\cdot)}{\partial z}$, $P(z)=(K)'/K$ and $Q(z)=1/(i K)$. Second, it is an easy computation to show that rewriting Eq. \ref{can_form} for $y = \frac{z }{\zeta}

@@

where $(\cdot)' = \frac{\partial (\cdot)}{\partial y}$, $\widetilde{P}(y)=\gamma'(y)/\gamma(y)$ and $\widetilde{Q}(y)=1/( i \cdot \zeta \gamma(y))$, with $\gamma(y)=y(y-1)^2$.

@@

	x^2+x+1/( i \zeta)=0.

@@

The model assumes $Pr = 1$, i.e. $K_M = K_H = K$, although in reality they are slightly different. This assumption is justified based on the degree of accuracy of the considered study. 
Further, as stated in the preceeding section, we assume eddy diffusivities in line with the classic O'Brien's type \citep{O'Brien1970}, viz. $K = Az(z-\zeta)^2$, where $A$ is a prefactor and $\zeta > H$ controls both shape and magnitude of $K$.
The O'Brien model has often been used in studies of stable boundary layers, see for instance \citet{pielke1984mesoscale} and \citet{Stull1988}, and a generalized O'Brien model was also recently adopted in \citet{grisogono2001katabatic} to study slope flows.
The properties of $K(z)$ in slope flows were thorougly justified in \citet{Grisogono2002};  we here recall the main criteria that $K(z)$ should honor:
%
\begin{enumerate}
\item realizability condition: $K(z) \ge 0$;
\item $K(z << \infty) = 0$;
\item $\max(K(z))$, reached at $z=h$, must lie within the dynamic and thermal boundary layers;
\item $h > \max(2 z_j, z_{inv})$;
\end{enumerate}
%
where $z_j$ is the height of the LLJ, and $z_{inv}$ is the height of the surface inversion layer (in general we have $z_{inv} > z_j$).
%
In addition, assuming Monin-Obukhov similarity theory to hold for the near wall regions (which is itself a subject of debate), one would ideally specify $K(z)$ in order to satisfy:
%
\begin{linenomath*}
\begin{subequations}
\begin{align}
   K(z) & = \kappa u_* z & z \in [z_0,z_s], \label{MOST1}\\
   K(z) & = f(u_*,z) \ | \ d^2f/dz^2 < 0 & z \in (z_s,H], \label{MOST2}
\end{align}
\end{subequations}
\end{linenomath*}
%%
%\begin{linenomath*}
%\begin{subequations}
%\begin{align}
%   K(z) & = \kappa u_* z/[1+\Phi(z)] & z \in [z_0,z_s], \label{MOST1}\\
%   K(z) & = f(u_*,z) \ | \ d^2f/dz^2 < 0 & z \in (z_s,H], \label{MOST2}
%\end{align}
%\end{subequations}
%\end{linenomath*}
%
where $z_s$ is the upper boundary of the constant-flux layer, $\kappa$ is the Von K\'arm\'an constant, and $u_* = \kappa z (du/dz)|_{z_0}$, is the so-called friction velocity.
Based on recent direct (DNS) and large-eddy simulation (LES) results (not shown) we make the additional assumption that $h$ is located in the neighborhood of the first zero of the $u$ velocity profile, well above $z_j$ but within the dynamic and thermal boundary layers (under such constraint $h$ identifies the lower boundary of the return flow). 
We will show hereafter that the solution is poorly sensitive on the exact value of $h$, as long as the resulting domain height $H<\zeta$ is large enough to accomodate the dynamic and thermal boundary layers.
Knowledge of $dK/dz|_{z=z_0}$ and $h$ allows to univocally specify $K(z)= Az(z-\zeta)^2$. A possible alternative would be to specify $dK/dz|_{z=z_0}$ and $\max{(K)}$. 
The size of the integration interval should be chosen as large as possible, under the constraint $H<\zeta$. Throughout the study we will set $H = \zeta-\zeta/100$, which allows to fully represent the dynamic and thermal boundary layers over a broad range of $z_0$ values.
Note that in general it is not possible to satisfy Eqs. \ref{MOST1} directly, since the specification of $K$ has to be done a-priori, and $u_*$ depends on the solution. 
The chosen normalization allows to overcome this limitation, and highlights the linear dependence of the length scales of the flow on $u_*$ ($A \propto u_* \propto L)$.
One could therefore first solve for $\overline{u}$ and $\overline{b}$, and then re-dimensionalize the problem, imposing $A=\kappa u_* / \zeta^2$ thus resulting in $K(z)=\kappa u_* z$, in agreement with Eq. \ref{MOST1}.

Despite the restrictive form of the model parameter $K(z)$, the formulation allows for exact integration of Eqs. \ref{eqn1u} and \ref{eqn1b}, thus providing a useful reference to test approximate solutions such as the WKB \citep{grisogono2001katabatic} and the dependence of the flow on the additional model parameter $z_0$. Further, the formulation can be used to validate numerical solutions of the Prandtl model equations (Eqs. \ref{eqn1u} and \ref{eqn1b}).

@@

      \includegraphics[width= 70.0mm]{./comparison_normalized_analytic_solutions_z0=1e-05}
      \caption{Comparison of the new analytic solution (A1) against the constant-$K$ (A2), the WKB (A3) and a numerical $1D$ solution (N1). Analytic profiles of normalized velocity $(\overline{u})$ are denoted with dot-dashed lines whereas analytic profiles of normalized buoyancy $(\overline{b})$ are denoted by dashed lines. Here $\overline{z}_0 = 0.00001$ and $\overline{\zeta} = 0.1$. A corresponding dimensional profile is characterized by $\max{(K)} = 0.1 \ \mathrm{m^2s^{-1}}$ and $K(z=0) = 10^{-5} \ \mathrm{m^2s^{-1}}$, in agreement with common atmospheric values.}

@@

In Fig. \ref{fig1} we compare the new analytic solution (A1 in the following) against the constant-$K$ (A2), the WKB (A3) and a numerical solution (N1), for a chosen $z_0,\zeta$. $\zeta$ is specified so that $h$ is located at the first zero of the $u(z)$ profile. 

@@

A1 shows a remarkably strong inversion in the near surface regions $(z^+ <1)$, when compared against its analytical counterparts, suggesting an overdiffusive behaviour of both A2 and A3. 
